---
title: "p8105_hw2_lg3239"
author: "Landi Guo"
date: "2022-10-04"
output: github_document
editor_options: 
  chunk_output_type: console
---
```{r, echo=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
```{r}
NYC_trans = 
  read_csv(
    "files/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%
  janitor::clean_names() %>%
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>%
  mutate(entry = recode(entry, 'YES' = TRUE, 'NO' = FALSE))
```
This dataset contains variables `r names(NYC_trans)`. It has `r nrow(NYC_trans)` rows and `r ncol(NYC_trans)` columns. The cleaning steps begin with importing the file and specifying the `route` columns all be character type. Then using `select()` to select the variables that we want. `mutate()` and `recode()` are used to transform `entry` from character type to logical values. However, this dataset is not tidy because the route numbers are supposed to be values of a variable route.

```{r}
NYC_trans %>%
  select(line, station_name) %>%
  distinct()
```
There are 465 distinct stations. 
`distinct()` gives all unique combinations between line and station names.

```{r}
NYC_trans %>%
  filter(ada == TRUE) %>%
  select(line, station_name) %>%
  distinct()
```
84 stations are ADA compliant
Use `filter()` to select the dataframe according to ADA compliance, then select the `line` and `station_name` to form distinct combinations.

```{r}
NYC_trans %>%
  filter(vending == 'NO') %>%
  pull(entry) %>%
  mean()
```
About 0.38 of station entrances/exits without vending allow entry.
Use `filter()` to select the dataframe according to vending availability, then choose the `entry` value since its logical, `mean()` is able to cast it as numeric.

```{r}
NYC_trans %>%
  pivot_longer(
    route1:route11,
    names_to = 'route_num',
    values_to = 'route') %>%
  filter(route == 'A') %>%
  select(line, station_name) %>%
  distinct()

NYC_trans %>%
  pivot_longer(
    route1:route11,
    names_to = 'route_num',
    values_to = 'route') %>%
  filter(route == 'A', ada == TRUE) %>%
  select(line, station_name) %>%
  distinct()
```
There are 60 distinct stations serve the A train, and 17 of them are ADA compliant.
`pivit_longer()` is used to make route number and route name as distinct variables, and `filter()` is used according to train A and ADA compliance. Similar to steps above, obtain the unique combinations between line and station names.

## Problem 2
```{r}
mr_trash_wheel = 
  readxl::read_excel("files/Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N549") %>%
  janitor::clean_names() %>%
  mutate(sports_balls = as.integer(round(sports_balls, digits = 0)))
mr_trash_wheel['trash_wheel'] = 'mr_trash'
```
`read_excel()` with specified `sheet` number and `range` is used to import the Mr.Trash Wheel dataset. For cleaning, `janitor::clean_names()` is used. `mutate()`, `round()`, and `as.integer()` are used to round the number of sports balls to the nearest integer. An addition variable `trash_wheel` with value `mr_trash` is added, specifying the dataset it belongs to.
\
This dataset contains `r names(mr_trash_wheel)`, with `r nrow(mr_trash_wheel)` rows and `r ncol(mr_trash_wheel)` columns. It has the information about the various trash collected according to date.

```{r}
prof_trash_wheel = 
  readxl::read_excel("files/Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M96") %>%
  janitor::clean_names() %>%
  mutate(year = as.character(year))
prof_trash_wheel['trash_wheel'] = 'prof_trash'
```
Similarly, `read_excel()` with specified `sheet` number and `range` is used to import the Professor Trash Wheel dataset. For cleaning, `janitor::clean_names()` is used. `mutate()` is used to conver `year` variable to character type as same as Mr.Trash Wheel. An addition variable `trash_wheel` with value `prof_trash` is added, specifying the dataset it belongs to.\
This dataset contains `r names(prof_trash_wheel)`, with `r nrow(prof_trash_wheel)` rows and `r ncol(prof_trash_wheel)` columns. It has the information about the various trash collected according to date. But it does not have the `sports_balls` variable as Mr.Trash Wheel.

```{r}
trash_wheel_df =
  full_join(mr_trash_wheel, prof_trash_wheel)

mr_trash_wheel %>%
  filter(year == '2020') %>%
  pull(sports_balls) %>%
  sum()

prof_trash_wheel %>%
  pull(weight_tons) %>%
  sum()
```
`full_join()` is used to merge the two datasets, keeping data that appear in either of the two.\
The resulting dataset has variables `r names(trash_wheel_df)`, and is `r nrow(trash_wheel_df)` x `r ncol(trash_wheel_df)`.

Using `prof_trash_wheel` on the `weight_tons` variable, and using `sum()`, the total weight of trash collected by Professor Trash Wheel is 190.12 tons.
Using `mr_trash_wheel`, `filter()` by 2020, on the `sports_balls` variable, and using `sum()`, the total number of sports balls collected by Mr. Trash Wheel in 2020 is 856.

## Problem 3
```{r}
pols_month_df = 
  read_csv("files/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(month = gsub("0", "", month)) %>%
  mutate(month = month.abb[as.numeric(month)]) %>%
  mutate(president = case_when(prez_gop == 1 ~ "gop",
                               prez_dem == 1 ~ "dem")) %>%
  select(-prez_gop, -prez_dem, -day)
```
`read_csv()` is used to import the pols_month dataset. `separate()` is used with `sep = "-"` to breakup `mon` into `year`, `month`, and `day`. Then `mutate()` is used to change the `month` to numeric values (without zero) and using `month.abb[]` to obtain month abbreviation. `mutate()` is also used with `case_when()`, specifying when `prez_gop` is 1, add the `president` variable with value `gop`, and when `prez_dem` is 1, add with value `dem`. Finally, remove the `prez_gop`, `prez_dem`, `day` columns by `select()`.\
This dataset contains `r nrow(pols_month_df)` observations of `r ncol(pols_month_df)` variables related to the number of national politicians who are democratic or republican at the time. The variables are `r names(pols_month_df)`. The year range is from `r pols_month_df %>% pull(year) %>% min()` to `r pols_month_df %>% pull(year) %>% max()`.

```{r}
snp_df = 
  read_csv("files/fivethirtyeight_datasets/snp.csv") %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(month = month.abb[as.numeric(month)]) %>%
  select(-day) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(year = if_else(year > 49, year + 1900, year + 2000)) %>%
  mutate(year = as.character(year)) %>%
  relocate(year) %>%
  arrange(year, month)
```
`read_csv()` is used to import the dataset. `separate()` is used with `sep = "/"` to breakup `date` into `month`, `day`, and `year`. Then `mutate()` is used to change the `month` to numeric values (without zero) and using `month.abb[]` to obtain month abbreviation. After removing `day` using `select()`, `mutate()` is used to convert year to numeric type, and with `if_else()` to convert year that is greater than 49 to four digits 19xx, otherwise to 20xx. `year` should be in character type. Finally, `relocate()` year to first column and `arrange()` according to year and month.\
This dataset contains `r nrow(snp_df)` observations of `r ncol(snp_df)` variables related to representative measure of stock market. The variables are `r names(snp_df)`, in which `close` is the closing values of the S&P stock index on the associated date. The year range is from `r snp_df %>% pull(year) %>% min()` to `r snp_df %>% pull(year) %>% max()`.

```{r}
unemploy_df = 
  read_csv("files/fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "rate"
  ) %>%
  janitor::clean_names() %>%
  mutate(year = as.character(year))
```
`read_csv()` is used to import the dataset. `pivit_longer()` is used on variables from `Jan` to `Dec` to cast the month and rate as distinct variables. `mutate()` changes `year` to character type.\
This dataset contains `r nrow(unemploy_df)` observations of `r ncol(unemploy_df)` variables. The variables are `r names(unemploy_df)`. The year range is from `r unemploy_df %>% pull(year) %>% min()` to `r unemploy_df %>% pull(year) %>% max()`.

```{r}
result_df = pols_month_df %>%
  left_join(snp_df) %>%
  left_join(unemploy_df)
```
Merge `snp_df` to `pols_month_df` using `left_join()`, and `unemploy_df` to the resulting dataset using `left_join()`.\
The final dataset contains `r nrow(result_df)` observations of `r ncol(result_df)` variables. The year range is from `r result_df %>% pull(year) %>% min()` to `r result_df %>% pull(year) %>% max()`. The variables are `r names(result_df)`.
